---
title: "STOR390_HW4"
output:
  html_document: default
  pdf_document: default
date: "2024-03-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This homework is designed to give you practice fitting a logistic regression and working with statistical/philosophical measures of fairness. We will work with the `titanic` dataset which we have previously seen in class in connection to decision trees.

Below I will preprocess the data precisely as we did in class. You can simply refer to `data_train` as your training data and `data_test` as your testing data.

```{r}

#this is all of the preprocessing done for the decision trees lecture.  

path <- 'https://raw.githubusercontent.com/guru99-edu/R-Programming/master/titanic_data.csv'
titanic <-read.csv(path)
head(titanic)

library(dplyr)

#replace ? with NA
replace_question_mark <- function(x) {
  if (is.character(x)) {
    x <- na_if(x, "?")
  }
  return(x)
}

titanic <- titanic %>%
  mutate_all(replace_question_mark)

set.seed(678)
shuffle_index <- sample(1:nrow(titanic))
head(shuffle_index)

titanic <- titanic[shuffle_index, ]
head(titanic)

library(dplyr)
# Drop variables
clean_titanic <- titanic %>%
select(-c(home.dest, cabin, name, x, ticket)) %>% 
#Convert to factor level
    mutate(pclass = factor(pclass, levels = c(1, 2, 3), labels = c('Upper', 'Middle', 'Lower')),
    survived = factor(survived, levels = c(0, 1), labels = c('No', 'Yes'))) %>%
na.omit()
#previously were characters
clean_titanic$age <- as.numeric(clean_titanic$age)
clean_titanic$fare <- as.numeric(clean_titanic$fare)
glimpse(clean_titanic)

create_train_test <- function(data, size = 0.8, train = TRUE) {
    n_row = nrow(data)
    total_row = size * n_row
    train_sample <- 1: total_row
    if (train == TRUE) {
        return (data[train_sample, ])
    } else {
        return (data[-train_sample, ])
    }
}
data_train <- create_train_test(clean_titanic, 0.8, train = TRUE)
data_test <- create_train_test(clean_titanic, 0.8, train = FALSE)

```

# 

Create a table reporting the proportion of people in the training set surviving the Titanic. Do the same for the testing set. Comment on whether the current training-testing partition looks suitable.

```{r}

train_survivors_prop <- mean(data_train$survived == "Yes")

test_survivors_prop <- mean(data_test$survived == "Yes")

survivors_table <- data.frame(
  Dataset = c("Training", "Testing"),
  Proportion_of_Survivors = c(train_survivors_prop, test_survivors_prop)
)

survivors_table
```

*student input*

# 

Use the `glm` command to build a logistic regression on the training partition. `survived` should be your response variable and `pclass`, `sex`, `age`, `sibsp`, and `parch` should be your response variables.

```{r}

logit_model <- glm(survived ~ pclass + sex + age + sibsp + parch, data = data_train, family = binomial)

summary(logit_model)

```

We would now like to test whether this classifier is *fair* across the sex subgroups. It was reported that women and children were prioritized on the life-boats and as a result survived the incident at a much higher rate. Let us see if our model is able to capture this fact.

# 

Subset your test data into a male group and a female group. Then, use the `predict` function on the male testing group to come up with predicted probabilities of surviving the Titanic for each male in the testing set. Do the same for the female testing group.

```{r}

male_test_data <- subset(data_test, sex == "male")
female_test_data <- subset(data_test, sex == "female")


male_predicted_probabilities <- predict(logit_model, newdata = male_test_data, type = "response")
female_predicted_probabilities <- predict(logit_model, newdata = female_test_data, type = "response")

mean(male_predicted_probabilities)
mean(female_predicted_probabilities)

```

# 

Now recall that for this logistic *regression* to be a true classifier, we need to pair it with a decision boundary. Use an `if-else` statement to translate any predicted probability in the male group greater than $0.5$ into `Yes` (as in Yes this individual is predicted to have survived). Likewise an predicted probability less than $0.5$ should be translated into a `No`.

Do this for the female testing group as well, and then create a confusion matrix for each of the male and female test set predictions. You can use the `confusionMatrix` command as seen in class to expidite this process as well as provide you necessary metrics for the following questions.

```{r}
library(caret)

male_test_data$survived <- factor(male_test_data$survived, levels = c("No", "Yes"))
female_test_data$survived <- factor(female_test_data$survived, levels = c("No", "Yes"))

male_predictions <- ifelse(male_predicted_probabilities > 0.5, "Yes", "No")
female_predictions <- ifelse(female_predicted_probabilities > 0.5, "Yes", "No")

male_confusion_matrix <- confusionMatrix(factor(male_predictions), reference = male_test_data$survived)
female_confusion_matrix <- confusionMatrix(factor(female_predictions), reference = female_test_data$survived)


male_confusion_matrix
female_confusion_matrix

```

# 

We can see that indeed, at least within the testing groups, women did seem to survive at a higher proportion than men (24.8% to 76.3% in the testing set). Print a summary of your trained model and interpret one of the fitted coefficients in light of the above disparity.

```{r}
summary(logit_model)
```

*The coefficient for the sexmale variable is -2.684206, which means that when all other variables are held constant, being male is associated with a decrease in the log odds of survival by approximately 2.684206 units compared to being female. This coefficient indicates that being male was a significant predictor of death on the Titanic and highlights the gender disparity in survival on the Titanic.*

# 

Now let's see if our model is *fair* across this explanatory variable. Calculate five measures (as defined in class) in this question: the Overall accuracy rate ratio between females and males, the disparate impact between females and males, the statistical parity between females and males, and the predictive equality as well as equal opportunity between females and males (collectively these last two comprise equalized odds). Set a reasonable $\epsilon$ each time and then comment on which (if any) of these five criteria are met.

```{r}

male_predictions <- ifelse(male_predicted_probabilities > 0.5, 1, 0)
female_predictions <- ifelse(female_predicted_probabilities > 0.5, 1, 0)

male_true_labels <- ifelse(male_test_data$survived == "Yes", 1, 0)
female_true_labels <- ifelse(female_test_data$survived == "Yes", 1, 0)

# Overall accuracy ratio
overall_accuracy_ratio <- sum(female_predictions == female_true_labels) / sum(male_predictions == male_true_labels)

# Disparate impact
disparate_impact <- sum(female_predictions) / sum(male_predictions)

# Statistical parity
statistical_parity <- abs((sum(female_predictions) / sum(male_predictions)) - 1)

# False Positive Rate (FPR) for males and females
male_fpr <- sum(male_predictions == 1 & male_true_labels == 0) / sum(male_true_labels == 0)
female_fpr <- sum(female_predictions == 1 & female_true_labels == 0) / sum(female_true_labels == 0)

# Predictive equality
predictive_equality <- abs(female_fpr - male_fpr)

# Equal opportunity
male_fnr <- sum(male_predictions == 0 & male_true_labels == 1) / sum(male_true_labels == 1)
female_fnr <- sum(female_predictions == 0 & female_true_labels == 1) / sum(female_true_labels == 1)
equal_opportunity <- abs(female_fnr - male_fnr)


overall_accuracy_ratio
disparate_impact
statistical_parity
predictive_equality
equal_opportunity


```

*The model does not meet the criteria for fairness. The disparate impact, statistical parity, and overall accuracy rate ratio suggest potential disparities in predictions between females and males. The predictive equality and equal opportunity measures show relatively lower disparities in false positive and false negative rates between the two groups, but it's not enough to reduce make the model fair.*.

It is always important for us to interpret our results in light of the original data and the context of the analysis. In this case, it is relevant that we are analyzing a historical event post-facto and any disparities across demographics identified are unlikely to be replicated. So even though our model fails numerous of the statistical fairness criteria, I would argue we need not worry that our model could be misused to perpetuate discrimination in the future. After all, this model is likely not being used to prescribe a preferred method of treatment in the future.

# 

Even so, provide a *philosophical* notion of justice or fairness that may have motivated the Titanic survivors to act as they did. Spell out what this philosophical notion or principle entails?

*Virtue ethics may have motivated survival outcomes on the Titanic because people believed it was a man's duty to protect his wife and children which meant sacrificing his own life so they would survive. The virtues of duty and chivalry were pretty universal, especially at that time, so women and children were saved first.*

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
